Tidal streams in triaxial systems

### 2015-02-08 ms.tex
words

% packages
\usepackage{amssymb,amsmath,amsbsy}

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\msun}{\mathrm{M}_\odot}
\newcommand{\bs}[1]{\boldsymbol{#1}}

@@

\section{Methods}\label{sec:methods}
\subsection{Potential form}

\subsection{Tidal disruption in action-angle coordinates}
When a star is tidally stripped from a progenitor system, the difference in angle coordinates, relative to the progenitor, grows linearly with time according to
\begin{equation}
	\Delta\bs{\theta} = \Delta\bs{\theta}(t=0) + \Delta\bs{\Omega}t
\end{equation}
where $\Delta\bs{\theta}(t=0)$ is the initial offset (at time of stripping, $t=0$) between the star and progenitor, and $\Delta\bs{\Omega}$ is the corresponding difference in fundamental frequencies. Once a star has ``entered'' the stream, the constant offset becomes small compared to the term that depends on time, so that
\begin{equation}
	\Delta\bs{\theta} \approx \Delta\bs{\Omega}t.
\end{equation}
If the spread of actions of the progenitor is small enough --- corresponding to a mass [or mass ratio?] of XXX --- the frequency vector for a given star may be approximated by a Taylor expansion about the frequency vector of the progenitor,
\begin{equation}
	\Omega_i \approx \Omega_{0,i} + \frac{\partial^2 H}{\partial J_i J_j}\Delta J_j + \mathcal{O}(\Delta J^2)
\end{equation}
where we make use of the Einstein summation convention in the above expression. To first order in the difference in actions, the difference in frequencies between a star and the progenitor is approximated by
\begin{equation}
	\Delta\Omega_i \approx \frac{\partial^2 H}{\partial J_i J_j}\Delta J_j = D_{ij}\Delta J_j.\label{eq:deltafreq}
\end{equation}
where $D_{ij}$ is the Hessian of the potential in action-space. For a thin stream to form, the Hessian must be dominated by a single large eigenvalue, where the corresponding eigenvector determines the direction along which the debris will spread in the angles. If two or all of the largest eigenvalues are comparable in size, the debris will not form a thin stream and the observability (projected surface density) of the debris structure will be diminished. We would like to be able to predict the morphology of a debris structure formed from a disrupting progenitor on a given orbit in a specified potential using this fact.

\subsection{Numerically computing the Hessian in action-space}
The Hessian can be computed for an orbit in a given potential provided with a numerical scheme for computing the actions, angles, and frequencies of an orbit in an arbitrary potential. This is a non-trivial task, and we use a custom implementation of the method presented in \citep{sanders14} to compute these quantities. The procedure for computing the Hessian is as follows:
\begin{enumerate}
	\item numerically integrate the orbit of interest in the test potential;
	\item fit a \emph{toy potential} \citep[see][]{sanders14} to the orbit, and create a 3D grid of actions in this toy potential;
	\item transform the actions and angles of the grid from the toy potential to ordinary phase-space;
	\item integrate the grid of initial conditions in the test potential;
	\item solve for actions, angles, and frequencies of this grid;
	\item use the (now deformed) grid in action-space to linearly interpolate a regular grid about the orbit of interest;
	\item use a finite-difference scheme to compute first derivatives of the frequency vectors ($D_{ij} = \partial \Omega_i / \partial J_j$).
\end{enumerate}
Once we compute the Hessian, we may solve for the eigenvalues and eigenvectors of the matrix.

An alternate approach uses the debris particles themselves to solve for the components of the Hessian at position (in action-space) of the progenitor orbit. Equation \ref{eq:deltafreq} demonstrates the relationship between the components of the relative frequency vector to the components of the relative action vector (through the Hessian) for a single particle. We now consider a case where we have a set of $K$ particles that we know should obey the equations
\begin{equation}
	\Delta\Omega_i^{(k)} = D_{ij}\Delta J_j^{(k)}.
\end{equation}
This is analogous to the classic problem in linear algebra of solving an overconstrained set of linear equations, $\mathbf{y} = \mathbf{A}\mathbf{x}$. Here, $\mathbf{y}$ is the length-$3K$ (column) vector of frequencies for each star, $\mathbf{x}$ is a length-9 (column) vector containing the ``unravelled'' values of the Hessian, and $\mathbf{A}$ is a $3K\times 9$ matrix containing the relative actions for all stars:
\begin{align}
	\mathbf{y} &= \left[ \begin{array}{c} 
					\Delta\Omega_1^{(1)} \\ \Delta\Omega_2^{(1)} \\ \Delta\Omega_3^{(1)} \\ \vdots \\ 
					\Delta\Omega_1^{(K)} \\ \Delta\Omega_2^{(K)} \\ \Delta\Omega_3^{(K)}
				\end{array} \right]\\
	\mathbf{A} &= \begin{bmatrix} 
					\Delta J_1^{(1)} & \Delta J_2^{(1)} & \Delta J_3^{(1)} & 0 & 0 & 0 & 0 & 0 & 0  \\
					0 & 0 & 0 & \Delta J_1^{(1)} & \Delta J_2^{(1)} & \Delta J_3^{(1)} & 0 & 0 & 0  \\
					0 & 0 & 0 & 0 & 0 & 0 & \Delta J_1^{(1)} & \Delta J_2^{(1)} & \Delta J_3^{(1)}  \\
					&&&&\vdots \\
					\Delta J_1^{(K)} & \Delta J_2^{(K)} & \Delta J_3^{(K)} & 0 & 0 & 0 & 0 & 0 & 0  \\
					0 & 0 & 0 & \Delta J_1^{(K)} & \Delta J_2^{(K)} & \Delta J_3^{(K)} & 0 & 0 & 0  \\
					0 & 0 & 0 & 0 & 0 & 0 & \Delta J_1^{(K)} & \Delta J_2^{(K)} & \Delta J_3^{(K)}  \\
			       \end{bmatrix}\\
	\mathbf{x} &= \left[ \begin{array}{c} D_{11} \\ D_{12} \\ \vdots \\ D_{33} \end{array} \right].
\end{align}
The parameter vector, $\mathbf{x}$, is found trivially with ordinary least squares.

% \begin{bmatrix} xz & xw \\ yz & yw \end{bmatrix}

\section{Experiments}\label{sec:experiments}

\subsection{Comparing Hessian solvers}

@@

>>>>

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{bbold}
\usepackage{arydshln}
\usepackage{algpseudocode}
\renewcommand\algorithmicthen{}
\renewcommand\algorithmicdo{}

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\msun}{\mathrm{M}_\odot}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ident}{\mathbb{1}}

\newcommand{\act}{J}
%\newcommand{\jac}{\bs{\rm J}}
\newcommand{\jac}{\bs{G}}

@@

For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (d.o.f.), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,...,\act_n)\label{eq:aafreq}
\end{align}
where $n=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}; each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit.

If galactic halos are triaxial, their gravitational potentials will almost certainly not be globally integrable. For a given halo, if the potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
\begin{equation}
	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
\end{equation}
then a large number of regular orbits will survive the perturbation (e.g., the KAM theorem; cite Arnold)\footnote{3-vectors are represented with bold symbols}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. The types of motion generic to typical triaxial systems can be split into several sub-categories (summarized in Figure~\ref{fig:orbit-tree}). Regular motion can be classified by the number of resonance relations obeyed \citep[e.g.,][]{lichtenberg83, valluri98}: \emph{conditionally periodic} orbits obey no resonance relations, \emph{uni-resonant} orbits obey a single relation of the form $\bs{m}\cdot\bs{\Omega}=0$, and \emph{bi-resonant} orbits obey two resonance relations, $\bs{m}\cdot\bs{\Omega}=0$ and $\bs{n}\cdot\bs{\Omega}=0$, where $\bs{m}$ and $\bs{n}$ are integer vectors.

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/orbit-tree.pdf}
\caption{} \label{fig:orbit-tree}
\end{center}
\end{figure*}

Resonant orbits are dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable resonant orbits generate families of orbits that behave or appear similarly, whereas unstable resonances typically generate regions of stochasticity. Not all orbits can be trivially associated with resonances, however, due to the infinitely complex, hierarchical structure of stable resonances. In autonomous, two d.o.f. systems, stochastic regions are generally surrounded and separated by regular motion, but in three or more d.o.f. the chaotic regions will connect and overlap; this is the origin of the phenomenon of Arnold diffusion \citep{arnold79}. The implication of Arnold diffusion is that chaotic orbits may radically change shape by means of traversing the web of unstable resonances and are strictly only bounded to their own energy hypersurface. Another way to understand this phenomenon is by means of orbital tori: in two dimensions, if a given orbital torus is destroyed due to perturbations, the orbit is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' confinement.

For small perturbations, generically, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed, then the uni-resonant tori are destroyed, and finally the bi-resonant tori -- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. The irregular or chaotic orbits that fill the parts of action-space where tori have been destroyed are loosely classified by being either \emph{strongly} or \emph{weakly} chaotic. Qualitatively, strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. There are many methods for classifying and measuring the degree of stochasticity of an orbit.

\section{Numerical methods}

\subsection{Potential choice}

The density distributions within dark matter halos formed in cosmological N-body simulations are typically triaxial (e.g., millenium, aquarius, via lactea). \citet{jing02} found that a triaxial generalization of the classic NFW density profile \citep{navarro96} generates excellent fits to their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of the many halos formed in their simulations. They find median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that \citet{jing02} use the opposite notation so that $c$ is the major and $a$ is the minor axis.} They find significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

All of these parameters are specified in terms of the \emph{density}; for orbit analysis, we need to determine the form of the potential in terms of these parameters, which, in general, requires numerical integration of the density at each position of interest. For computational efficiency, many authors instead express the triaxiality in the form of the potential, however this can lead to unphysical situations where the density becomes negative. \citet{lee03} derive a perturbative expansion of the potential integral and show that the expansion is accurate even for modest axis ratios (e.g., the median values shown above).

For this work, we use the triaxial potential expression from \citet{lee03}, parametrized in a slightly different manner. In terms of spherical coordinates\footnote{(radius, azimuth, colatitude)} with the radius normalized by the scale radius, $u = r/r_s$
\begin{align}
	\Phi(u,\phi,\theta) \approx \frac{v_c^2}{\ln2 - 1/2}\left[F_1(u) + \frac{1}{2}(e_b^2 + e_c^2)F_2(u) + \frac{1}{2} [(e_b\sin\theta \sin\phi)^2 + (e_c\cos\theta)^2] F_3(u) \right]
\end{align}
where $v_c$ is the circular velocity at the scale radius, $r_s$ (for the spherical case), $e_b = \sqrt{1 - (b/a)^2}$, and $e_c = \sqrt{1 - (c/a)^2}$. The functions $F_i(u)$ are given in the appendix of \cite{lee03}. We chose $r_s=20~{\rm kpc}$ and $v_c \approx 200~{\rm km}~{\rm s}^{-1}$ , which [...] (Figure~\ref{fig:potential}).

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/potential.pdf}
\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:orbit-tree}
\end{center}
\end{figure*}

[see discussion section for adding terms (disk, time evolution) to potential]

\subsection{Orbit integration}

[Use a blah blah integrator...]

\subsection{Lyapunov exponents}

The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent of an orbit. Defining the vector $\bs{w} = (q_1,...,q_N,p_1,...,p_N)$ for some set of canonical coordinates $(q_i,p_i)$, we can write Hamilton's equations as 
\begin{equation}
	\dot{\bs{w}} = \bs{\mathcal{J}} \frac{\partial H}{\partial \bs{w}} = f(\bs{w}) \label{eq:ham}
\end{equation}
where $\bs{\mathcal{J}}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
\begin{equation}
	\bs{\mathcal{J}} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
\end{equation}
with $N$-dimensional identity matrices $\ident$. If we consider a nearby position separated from $\bs{w}$ by an infinitesimal deviation $\delta \bs{w}$, such that $\bs{w}' = \bs{w} + \delta \bs{w}$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as follows
\begin{align}
	\dot{\bs{w}}' &= f(\bs{w}')\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &= f(\bs{w} +  \delta \bs{w})\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &\approx f(\bs{w}) + \jac_{\bs{w}}\cdot\delta \bs{w} + \mathcal{O}(\delta \bs{w}^2)\\
	\dot{\delta \bs{w}} &\approx \jac_{\bs{w}} \cdot  \delta \bs{w}\label{eq:deviate}
\end{align}
where $\jac$ is the Jacobian of Eq.~\ref{eq:ham}, evaluated at the parent orbit. For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
\begin{algorithmic}[1]
\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
\State {\bf define} renormalization integration period, $\tau \ll hK$ 
\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
\If {a normalization timestep}
\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
\EndIf
\EndFor 
\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}
[Some more practical issues with implementation] [Figure showing $l_N$ for regular, chaotic orbit?]

[Copy in text fomr ``behavior of MLE'']

[Computationally, expensive to compute Lyapunov exponent for sticky or mildly chaotic orbits]

\subsection{Numerical estimation of fundamental frequencies}

Bounded, regular orbits in triaxial potentials will have three \emph{fundamental frequencies}, $\bs{\Omega}$, that determine the periodic behavior of motion (Eq.~\ref{eq:aafreq}).

[Figure: reproduce some result?]

or
[Figure: for same orbits as in Lyapunov section, show frequency diffusion?]

\subsection{Particle balls}

Need a fast way to characterize morphology of local orbits after finite time

Can't run N-body for many tens of thousands of initial conditions

Explain simplistic approach

[Figure: Compare N-body simulation to balls of particles -- captures the essence]

Cite papers that use Lagrange-point stripping methods, Rewinder paper

Like only considering debris that comes off at a single pericentric passage

\section{Experiments}

\subsection{Frequency map}

We generate frequency maps at X energies for tube orbits started around the major and intermediate axes of the potential (figures)

Explain qualitative differences between the two -- e.g., Y-Z plane will have many more stochastic tubes because intermediate axis

\subsection{Morphology}

\subsubsection{Entropy evolution}

\subsubsection{Density evolution}

Compare mean density or some other metric for the balls, show that decreases faster for chaotic orbit -- chaotic mixing dominates over phase-mixing

Power law (Helmi and White)

\subsubsection{Adding a disk?}

@@

\subsection{Potential}

Time dependence

Radial dependence of axis ratios

Why not disk + bulge? What do we expect when we add? See later section for test of this...
